

\beginedxvertical{Page One}

\beginedxtext{Preliminaries}





At the end of this sequence, and after some practice, you should be able to:

\begin{itemize}
\item Compute the least-squares solution to matrix-vector equations.
\item Apply the least-squares technique to 
\end{itemize}

For time budgeting purposes, this sequence has 3 videos totaling 14 minutes, 
plus some questions.  



\endedxtext

\endedxvertical



\beginedxvertical{Introducing the Question}


\beginedxproblem{Matrix-Vector Solutions}{\dpa1}

Let $A$ be a  real  $m\times n$ matrix.  The set of vectors $v$ for which the equation $Ax=v$ has a solution...

\edXabox{type="multichoice" expect="is some subspace of $\R^m$ but not necessarily all of it" options="is all of $\R^m$","is some subspace of $\R^m$ but not necessarily all of it","is a subset of $\R^m$ but not necessarily a subspace of it"}


\edXsolution{ 
The equation $Ax = v$ has a solution if and only if $v$ is in the span of the columns of $A$.  The span of 
the columns of $A$ might not be all of $\R^m$, but it is a subspace of $\R^m$.  
}


\endedxproblem

\doedxvideo{What is Least-Squares?}{RLHQocdvsZs}


\endedxvertical





\beginedxvertical{Finding the Least-Squares Solution}

\doedxvideo{Finding the Least-Squares Solution}{B_-rVbHJSSc}


\beginedxtext{Least-Squares Theorem}


{\keya{\bf{Theorem.}}}  
Let $A$ be an $m\times n$ matrix and let $v\in \R^m$.  Then the equation
$A^t Ax = A^t v$ is consistent, and any of its solutions will be a least-squares solution 
to the equation $Ax = v$.  

\endedxtext



\beginedxproblem{Least-Squares Practice}{\dpa3}

Let $A = \left[\begin{array}{cc} 1 & 1 \\ 1 & 0 \\ -1 & -1 \\ 0 & -1 \end{array} \right].$
Find the least squares solution to 
\[ Ax = \left[\begin{array}{c} 3 & -1 & 1 & 2 \end{array} \right]. \]

\input{vectorentry.tex}


\edXabox{type="custom" cfn="VectorEntry" expect="[[1.4],[-1.6]]"}

\edXsolution{ 
}

\endedxproblem




\endedxvertical





\beginedxvertical{Least-Squares Regression}


\beginedxproblem{Find a Parabola?}{\dpa1}

Consider the points $(0,1), (1,1), (2,2),$ and $(3,5)$.  

Is there an equation $y = cx^2 + ax + b,$ with $c,a,b$ all real, satisfied by all four points?  

\edXabox{expect="No" options="Yes","No"}


\edXsolution{
For a set of parameters $c,a,b$ to work, we would need the following system of equations to
be satisfied:
\[
\begin{array}{rcl}
1 & = &c\cdot 0^2 + a\cdot 0 + b \\
1 & = &c\cdot 1^2 + a\cdot 1 + b \\
2 & = &c\cdot 2^2 + a\cdot 2 + b \\
5 & = &c\cdot 3^2 + a\cdot 3 + b 
\end{array}
\]
This is a system of linear equations in $c,a,b$ given by the augmented matrix
\[
\left[\begin{array}{ccc:c}
0 & 0 & 1 & 1 \\
1 & 1 & 1 & 1 \\
4 & 2 & 1 & 2 \\
9 & 3 & 1 & 5
\end{array} \right]. \]
Row-reducing this matrix reveals that the system is inconsistent: there is no such equation
satisfied by all four points.  
}

\endedxproblem


\doedxvideo{Least-Squares Regression}{Aip0artsTD8}


\beginedxproblem{Closest Parabola}{\dpa3}

Finish the problem from above (and from the video) by finding the least-squares parabola of the form
$y = cx^2 + ax + b$ for the
data points $(0,1), (1,1), (2,2),$ and $(3,5)$.  


Enter the correct formula in the form $c*x^2 + a*x + b$, either exactly or to two decimal places.  (Remember to use * for multiplication.)    

\edXinline{$y=$ }
\edXabox{type="formula" expect="0.75*x^2-0.95*x+1.05" samples="x@1:5#5" feqin="1" inline="1" tolerance=".01"}

\edXsolution{
}

\endedxproblem




\endedxvertical





\beginedxvertical{Least-Squares Questions}


\beginedxproblem{Closest Parabola}{\dpa3}

Finish the problem from the previous page (and from the video) by finding the least-squares parabola of the form
$y = cx^2 + ax + b$ for the
data points $(0,1), (1,1), (2,2),$ and $(3,5)$.  


Enter the correct formula in the form $c*x^2 + a*x + b$, either exactly or to two decimal places.  (Remember to use * for multiplication.)    

\edXinline{$y=$ }
\edXabox{type="formula" expect="0.75*x^2-0.95*x+1.05" samples="x@1:5#5" feqin="1" inline="1" tolerance=".01"}

\edXsolution{
}

\endedxproblem



\beginedxproblem{Sine Function}{\dpa5}

We wish to find the function of the form $y = a + b\sin x$ that is the least-squares fit for the  data points $(0,3), (\frac{\pi}{2},1), (\pi, 5), (\frac{3\pi}{2},6),(2\pi, 2).$

This is equivalent to finding the least-squares solution to a matrix-vector equation
$
A\left[ \begin{array}{c} a \\ b\end{array} \right] = v.$

What is $A$?  (Make sure your rows are in order, corresponding to the order of the given data points.)  

\input{matrixentry.tex}

\edXabox{type="custom" cfn="MatrixEntry" expect="[[1,0],[1,1],[1,0],[1,-1],[1,0]]"}

What is $v$?  


\input{vectorentry.tex}

\edXabox{type="custom" cfn="VectorEntry" expect="[[3],[1],[5],[6],[2]]"}




\edXsolution{
}

\endedxproblem



\beginedxproblem{Sine Function Continued}{\dpa3}

Finish the analysis.  Find $a$ and $b$ giving the least-squares solution for the previous question.  

\edXinline{$a=$ }
\edXabox{type="numerical" expect="3.4" inline="1" tolerance=".01"}


\edXinline{$b=$ }
\edXabox{type="numerical" expect="-2.5" inline="1" tolerance=".01"}



\edXsolution{
}

\endedxproblem


\endedxvertical




